Formalizing Commutative Algebra in Coq: \\ Nakayama's Lemma\thanks{Source code for this work is available on the following site: \url{https://github.com/ku-sldg/algebra}}

### 2023-01-03 nakayama/paper.tex
Broke section on algebraic structures and formal proof into two separate sections

\paragraph{Commutative rings.}
In abstract algebra, the quintessential example of a commutative ring is the
set of integers
\[\mathbb{Z} = \{ \ldots, -3, -2, -1, 0, 1, 2, 3, \ldots \}.\]
using the natural definitions of addition and multiplication.

@@

\paragraph{Ideals of commutative rings.}
The concept of an {ideal} of a ring can be thought of as an extension of the
notion of an integer $n$ in the ring of integers $\mathbb{Z}$. An \emph{ideal}
of commutative ring $R$ is a subset $I$ of $R$ that is itself an abelian group
under addition, which also satisfies the following ``absorption'' property:
Given any element $a$ of $I$, the product  $x \cdot a$ is again in $I$ for any
ring element $x \in R$.

@@

\paragraph{Local rings.}
A commutative ring is \emph{local} if it has exactly one maximal ideal. 
Every field is local since the only proper ideal of a field is the zero ideal,
though by our observations above, the ring of integers is not local. 
However, the set of all rational numbers that can be written with an odd
denominator does form a subring of all rational numbers, and its unique maximal
ideal consists of the elements with even numerator; in fact, this ring is the
so-called \emph{localization} of $\mathbb{Z}$ at the maximal ideal
$2\mathbb{Z}$. The ring of integers modulo $n>1$ is local if and only if $n$ is
a power of a prime number $p$, in which case the unique maximal ideal consists
of all multiples of $p$.

@@

\paragraph{Modules over commutative rings.}
Let  $R$ be a commutative ring. 
A \emph{module over $R$}, or \emph{$R$-module}, is an abelian group $M$ under a
binary operation $+$, and a {scalar multiplication} $R \times M \to M$ denoted
$\cdot$, satisfying the following properties for all $r, s \in R$
and $u, v \in M$.

@@

\section{Formalization}
We start by describing our process of formalizing the necessary algebraic
structures detailed in the previous section. Then, with that in hand, we move
on to the formal proof of Nakayama's Lemma.

\subsection{Our Algebraic Hierarchy}
\begin{figure}[t]
  \caption{The hierarchy of our algebraic structures}
  \includegraphics[width=.4\textwidth]{algebraic-structures.pdf}
  \centering
\end{figure}

\paragraph{Semigroups.} Our foundation begins by defining a semigroup class, which declares a binary
operation to be associative. From here, we build up through monoids, which
introduce identities, to groups, which introduce inverses. Note the double
equals ``$==$'' appearing in these definitions is notation for an arbitrary
equivalence relation over the group's carrier set, which acts as equality.

@@

\paragraph{Algebraic quotients.}
While in the end, our formal proof does not call upon quotients of
algebraic structures, alternative proofs of Nakamaya's Lemma take advantage of 
 the structures quotient rings and quotient modules.  As algebraic 
 quotient structures are also fundamental to commutative algebra,
it is work pointing out that we have formalized quotients of groups, rings, and modules 
  using typeclasses, which appear to work rather smoothly.

@@

\paragraph{Rings and their ideals.}
Moving onward, rings form the next step in our algebraic hierarchy; a ring has
two binary operations: addition, which must be commutative, and multiplication,
which need not be commutative in general. In our formulation, rings must have a
multiplicative identity. Next, we formalized the definition of a commutative
ring, further requiring commutativity of multiplication. At this point, we
formalized the notion of an ideal of a commutative ring, a subgroup of the ring
under addition that satisfies the absorption property under multiplication,
i.e., \(r a\) sits in the ideal whenever \(a\) of the ideal, regardless of the choice of 
element \(r\) of the commutative ring. We also used this to formalize the
notion of a quotient ring $R/I$, where $I$ is an ideal of a commutative ring
$R$.

@@

%\emily[inline]{Emily:  Unfortunately, I think we need to update naming
%  conventions since in mathematics, ``vector'' is reserved for an element of a
%  vector space over a field, rather than an element of an arbitrary module.  So
%  this could cause confusion.  Similar with ``basis,'' which is especially
%  tricking since two different generating sets of a finitely generated module
%  can have different sizes, but all bases of a vector space have the same
%  cardinality, and an element of a module can typically be written as a scalar
%  combination in multiple ways, while bases are linearly independent.  I've
%  proposed a rewrite of the blue paragraph right above it.  Drew, can you
%  double check it for accuracy with respect to the code? 
%}
%\emily[inline]{We'll also need to update the code.  I think {\ttfamily vectors}
%  should be something like {\ttfamily scalar-combinations}, and
%  {\ttfamily basis} something like {\ttfamily generating-set}. 
%}

\paragraph{Modules over rings.}
At this point, we formalized the definition of a module over a commutative ring,
the commutative-algebraic generalization of the notion of a vector space over a
field. Nakayama's Lemma is a statement about finitely generated modules, and
hence we formalized the notion of a scalar combination of a finite
collection of elements, $u_1, \ldots, u_n$ a module $M$ over a commutative ring $R$, i.e.,
expressions of the form $r_1 u_1 + r_1 u_2 + \cdots + r_n u_n$, where each
$r_i$ is an element of $R$.

@@

\subsection{Constructing the Formal Proof}
Beyond formalizing the relevant structures from higher algebra, we also
formally establish some basic theory from the field in Coq. For instance, 
consider the notion of a \emph{unit} of a
commutative ring $R$, which is an element $a \in R$ with a multiplicative inverse,
i.e., there exists an element $a^{-1} \in R$ for which $x^{-1} \cdot x$ is the
multiplicative identity $1 \in R$.  In fact, it $x$ is a unit, its inverse is unique.

@@

\paragraph{Dealing with the Axiom of Choice.}
Every non-unit element of a commutative ring is contained in some maximal
ideal, a fact that relies on the Axiom of Choice. 
The following standard informal
proof calls upon Zorn's lemma, which is equivalent to the Axiom of Choice, and which 
says that if every chain of elements from a partially ordered set  has an upper
bound in the set, then the set must contain at least one maximal element.

@@

  Continuing in this way, if  this process terminates, we have successfully identified a
    maximal ideal of $R$ that contains $x$, and if not, one obtains an infinite, strictly increasing chain of proper ideals containing $x$, 
    \[I_{1}\subsetneq I_{2}\subsetneq I_{3}\subsetneq\cdots\subsetneq I_n \subsetneq \cdots. \]
    It is straightforward to verify that an increasing union of ideals 
    is, by definition, an ideal containing every ideal in the union.  Moreover, if each ideal in the union
    contains an element $x$, then the union certainly 
       contains \(x\) as well.

@@

To start describing our proof, we formalize the statement of Nakayama's Lemma
in Coq.

@@

\emily[inline]{Emily: Reflecting, I think it would be valuable to include at
least parts of the formal proof.  Pieces would work, though it is short, and it
could be ``broken up'' among the description of the informal argument.  As is,
our description of the proof focuses on the informal argument, which is not
``new''.}

>>>>

\subsubsection{Commutative rings}
In abstract algebra, the quintessential example of a commutative ring is the
set of integers
\[\mathbb{Z} = \{ \ldots, -3, -2, -1, 0, 1, 2, 3, \ldots \}.\]
using the natural definitions of addition and multiplication.

@@

\subsubsection{Ideals of commutative rings}
The concept of an {ideal} of a ring can be thought of as an extension of the
notion of an integer $n$ in the ring of integers $\mathbb{Z}$. An \emph{ideal}
of commutative ring $R$ is a subset $I$ of $R$ that is itself an abelian group
under addition, which also satisfies the following ``absorption'' property:
Given any element $a$ of $I$, the product  $x \cdot a$ is again in $I$ for any
ring element $x \in R$.

@@

\subsubsection{Local rings}
A commutative ring is \emph{local} if it has exactly one maximal ideal. 
Every field is local since the only proper ideal of a field is the zero ideal,
though by our observations above, the ring of integers is not local. 
However, the set of all rational numbers that can be written with an odd
denominator does form a subring of all rational numbers, and its unique maximal
ideal consists of the elements with even numerator; in fact, this ring is the
so-called \emph{localization} of $\mathbb{Z}$ at the maximal ideal
$2\mathbb{Z}$. The ring of integers modulo $n>1$ is local if and only if $n$ is
a power of a prime number $p$, in which case the unique maximal ideal consists
of all multiples of $p$.

@@

\subsubsection{Modules over commutative rings}
Let  $R$ be a commutative ring. 
A \emph{module over $R$}, or \emph{$R$-module}, is an abelian group $M$ under a
binary operation $+$, and a {scalar multiplication} $R \times M \to M$ denoted
$\cdot$, satisfying the following properties for all $r, s \in R$
and $u, v \in M$.

@@

\section{Our Algebraic Hierarchy}
In this section we describe our process of formalizing the necessary algebraic
structures, detailed in the previous section, in Coq. Then, with these formal definitions 
 in hand, we move
on in the next section to detail our formal proof of Nakayama's Lemma.

\begin{figure}[t]
  \caption{The hierarchy of our algebraic structures}
  \includegraphics[width=.4\textwidth]{algebraic-structures.pdf}
  \centering
\end{figure}

\subsection{Semigroups} Our foundation begins by defining a semigroup class, which declares a binary
operation to be associative. From here, we build up through monoids, which
introduce identities, to groups, which introduce inverses. Note the double
equals ``$==$'' appearing in these definitions is notation for an arbitrary
equivalence relation over the group's carrier set, which acts as equality.

@@

\subsection{Algebraic quotients}
While in the end, our formal proof does not call upon quotients of
algebraic structures, alternative proofs of Nakamaya's Lemma take advantage of 
 the structures quotient rings and quotient modules.  As algebraic 
 quotient structures are also fundamental to commutative algebra,
it is work pointing out that we have formalized quotients of groups, rings, and modules 
  using typeclasses, which appear to work rather smoothly.

@@

\subsection{Rings and their ideals}
Moving onward, rings form the next step in our algebraic hierarchy; a ring has
two binary operations: addition, which must be commutative, and multiplication,
which need not be commutative in general. In our formulation, rings must have a
multiplicative identity. Next, we formalized the definition of a commutative
ring, further requiring commutativity of multiplication. At this point, we
formalized the notion of an ideal of a commutative ring, a subgroup of the ring
under addition that satisfies the absorption property under multiplication,
i.e., \(r a\) sits in the ideal whenever \(a\) of the ideal, regardless of the choice of 
element \(r\) of the commutative ring. We also used this to formalize the
notion of a quotient ring $R/I$, where $I$ is an ideal of a commutative ring
$R$.

@@

%\emily[inline]{Emily:  Unfortunately, I think we need to update naming
%  conventions since in mathematics, ``vector'' is reserved for an element of a
%  vector space over a field, rather than an element of an arbitrary module.  So
%  this could cause confusion.  Similar with ``basis,'' which is especially
%  tricking since two different generating sets of a finitely generated module
%  can have different sizes, but all bases of a vector space have the same
%  cardinality, and an element of a module can typically be written as a scalar
%  combination in multiple ways, while bases are linearly independent.  I've
%  proposed a rewrite of the blue subsubsection right above it.  Drew, can you
%  double check it for accuracy with respect to the code? 
%}
%\emily[inline]{We'll also need to update the code.  I think {\ttfamily vectors}
%  should be something like {\ttfamily scalar-combinations}, and
%  {\ttfamily basis} something like {\ttfamily generating-set}. 
%}

\subsection{Modules over rings}
At this point, we formalized the definition of a module over a commutative ring,
the commutative-algebraic generalization of the notion of a vector space over a
field. Nakayama's Lemma is a statement about finitely generated modules, and
hence we formalized the notion of a scalar combination of a finite
collection of elements, $u_1, \ldots, u_n$ a module $M$ over a commutative ring $R$, i.e.,
expressions of the form $r_1 u_1 + r_1 u_2 + \cdots + r_n u_n$, where each
$r_i$ is an element of $R$.

@@

\section{Constructing the Formal Proof}
Beyond formalizing the relevant structures from higher algebra, we also
formally establish some basic theory from the field in Coq. For instance, 
consider the notion of a \emph{unit} of a
commutative ring $R$, which is an element $a \in R$ with a multiplicative inverse,
i.e., there exists an element $a^{-1} \in R$ for which $x^{-1} \cdot x$ is the
multiplicative identity $1 \in R$.  In fact, it $x$ is a unit, its inverse is unique.

@@

\subsection{Dealing with the Axiom of Choice}
Every non-unit element of a commutative ring is contained in some maximal
ideal, a fact that relies on the Axiom of Choice. 
The following standard informal
proof calls upon Zorn's lemma, which is equivalent to the Axiom of Choice, and which 
says that if every chain of elements from a partially ordered set  has an upper
bound in the set, then the set must contain at least one maximal element.

@@

  Continuing in this way, if  this process terminates, we have successfully identified a
    maximal ideal of $R$ that contains $x$, and if not, one obtains an infinite, strictly increasing chain of proper ideals of $R$, each containing $x$, 
    \[I_{1}\subsetneq I_{2}\subsetneq I_{3}\subsetneq\cdots\subsetneq I_n \subsetneq \cdots. \]
    It is straightforward to verify that an increasing union of ideals 
    is, by definition, an ideal containing every ideal in the union.  Moreover, if each ideal in the union
    contains an element $x$, then the union certainly 
       contains \(x\) as well.

@@

\subsection{The Formal Statement of Nakayama's Lemma}
To start describing our proof, we formalize the statement of Nakayama's Lemma
in Coq.

@@
